Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-20 doc/asplos13/grappa.tex

What makes Grappa's tasks special is that when they perform
long-latency operations they can give up control of their core,
allowing the core to remain busy while waiting for the operation to
complete. This most often done implicitly inside a call into Grappa's
API, but it can also be done explicitly by the programmer using the
calls shown in Figure~\ref{fig:scheduling}.

@@

Grappa provides multiple methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. The first is explicit task
spawns. When the programmer identifies work that can be done in
parallel, the work may be wrapped up in a function and queued with its
arguments for later execution using a \texttt{spawn}. Sometimes the
programmer may want to spawn a task on a specific core in the system,
or at the home core of a particular memory location. Grappa provides a
\texttt{spawn\_on} call for this purpose; the task is bound the
specified core immediately and will never be stolen.

The other method for expressing parallelism is a Cilk-like {\em
  recursive loop decomposition} \TODO{cite}. The programmer specifies
a function pointer along with start and end indices and an optional
threshold; Grappa decomposes the iterations into a
logarithmically-sized tree of tasks, stopping to execute when the
number of iterations is below the threshold.

@@

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      execution resources
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

@@

Large, low-locality data is stored in Grappa's global memory. All
global data must be accessed through calls into Grappa's API, shown in
Figure~\ref{fig:accessing-memory}.

Grappa provides two methods for storing data in the global memory. The
first is a distributed heap striped across all the machines in the
system. The \texttt{global\_malloc} and \texttt{global\_free} calls
are used to allocate and deallocate memory in the global heap; on
allocation, a global pointer is returned. Grappa also allows any local
data on the core's stacks or heap to be exported to the global address
space and accessed by other cores across the system.

There are two approaches to accessing global memory. When the
programmer expects global data to have spatial locality to exploit,
{\em cache} operations may be used. When there is no locality to
exploit, {\em delegate} operations are used. Since these operations
are expected communicate with other nodes and have high latency,

With a cache, applications can instruct Grappa to fetch a global
pointer of any length and return a local pointer to a cached copy of
the global memory. Under the hood, Grappa performs the mechanics of
gathering chunks of data from multiple system nodes and presenting a
conventional appearing linear block of memory as a pointer into a
cache. Grappa cache operations have the usual read-only and read-write
variants, along with a write-only variant used to initialize data
structures.

When accessing and modifying low-locality data, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).

Delegate operations are also the primary synchronization method in
Grappa. Each piece of global memory is paired with a single core in
the system; all accesses to that memory are done by that core. This
allows delegate operations to be performed atomically on the core
simply by making sure no other code interleaves; since context
switches in Grappa are cooperative, this is trivial. Thus, delegate
operations are able to provide atomic semantics without using atomic
operations.

@@

The first is a load balancing mechanism that uses work stealing to
move tasks from busy nodes to idle nodes. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the Garnet library. Most of our requests are small, and commodity
networks do not transport small packets efficiently. In order to make
more efficient use of bandwidth, Grappa aggregates packets going to the
same destination until it can form a larger packet that will make more
efficient use of the network. In this way, Grappa trades per-request
latency for application-level throughput.

>>>>

What makes Grappa's tasks special is that when they perform
long-latency operations they can give up control of their core,
allowing the core to remain busy while waiting for the operation to
complete. This most often done implicitly inside a call into Grappa's
API, but it can also be done explicitly by the programmer using the
calls shown in Figure~\ref{fig:scheduling}. \TODO{cite UPC split phase
    reads and writes, but include that we provide a way to overlap
    computation}

@@

Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. The first is explicit task
spawns. When the programmer identifies work that can be done in
parallel, the work may be wrapped up in a function and queued with its
arguments for later execution using a \texttt{spawn}. Sometimes the
programmer may want to spawn a task on a specific core in the system,
or at the home core of a particular memory location. Grappa provides a
\texttt{spawn\_on} call for this purpose; the task is immediately
enqueued on the specified core and is not available to other cores.

The next method for expressing parallelism is a parallel for loop,
where the number of iterations must be known at loop entry. The programmer specifies
a function pointer along with start and end indices and an optional
threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations,
 as in the implementation of Cilk's cilk\_for construct \TODO{cite cilk\_for implementation:
     could only find slides page6 of http://www.clear.rice.edu/comp422/lecture-notes/comp422-2012-Lecture5-Cilk++.pdf}.
It generates a logarithmically-deep tree of tasks, stopping to execute
the loop body when the number of iterations is below the threshold (``grainsize'' in Cilk).
\TODO{we may want to make our classification purer: the for loop is
    not Grappa's core, and is built atop tasks as a useful construct; in some sense the only way
to express parallelism in Grappa is tasks and perhaps call\_on. -BM}

@@

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism} % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

@@

Large data that is expected to be accessed with low locality is stored in Grappa's global memory. All
global data must be accessed through calls into Grappa's API, shown in
Figure~\ref{fig:accessing-memory}.

Grappa provides two methods for {\emph storing} data in the global memory. The
first is a distributed heap striped across all the machines in the
system. The \texttt{global\_malloc} and \texttt{global\_free} calls
are used to allocate and deallocate memory in the global heap; on
allocation, a global pointer is returned. Grappa also allows any local
data on a core's stacks or heap to be exported to the global address
space to be made accessible to other cores across the system.

There are also two approaches to {\emph accessing} global memory. When the
programmer expects a computation on shared data to have spatial
locality to exploit, {\em cache} operations may be used. When there is no locality to
exploit, {\em delegate} operations are used. Since these operations
are expected communicate with other nodes and have high latency,
\TODO{...}

With a cache, applications can instruct Grappa to fetch a global
pointer of any length and return a local pointer to a cached copy of
the global memory. Under the hood, Grappa performs the mechanics of
gathering chunks of data from multiple system nodes and presenting a
conventional appearing linear block of memory as a pointer into a
cache. Grappa cache operations have the usual read-only and read-write
variants, along with a write-only variant used to initialize data
structures. Cache operations exploit spatial locality by reducing the number of small network
messages accessing contguous data. Languages for distributed
shared memory systems have done optimizations to achieve a similar goal. The UPC compiler
coalesces struct and array accesses into 
remote get/put \cite{Chen:2005}, or Fortran D compiler's message
vectorization that hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a
mechanism for exploiting temporal locality by operating on the data locally.

When the access pattern has low-locality, it is more efficient
to modify the data on its home core rather than bringing a copy to the
requesting core and returning it after modification. Delegate
operations provide this capability. Applications can dispatch
computation to be performed on individual machine-word sized chunks of
global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).

Delegate operations, proposed in \cite{Nelson:hotpar11}, are also the primary synchronization method in
Grappa. Each piece of global memory is paired with a single core in
the system; all accesses to that memory are done by that core. This
allows delegate operations to be performed atomically on the core
simply by making sure no other code interleaves; since context
switches in Grappa are cooperative, this is trivial. Thus, delegate
operations are able to provide atomic semantics to memory owned by one core without using atomic
operations. Using delegation to implement isolation for any subset of
memory has been explored in \cite{delegated:oopsla11}.

@@

The first is a dynamic load balancing mechanism that uses work stealing to
move tasks from busy nodes to idle nodes. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. Most of \checkme{our requests} are small, and commodity
networks do not transport small packets efficiently. In order to make
more efficient use of bandwidth, Grappa aggregates packets going to the
same destination until it can form a larger packet that will make more
efficient use of the network. In this way, Grappa trades per-request
latency for application-level throughput.

### 2012-07-20 doc/asplos13/method.tex

\section{Methodolody} \label{sec:method}

@@

>>>>

\section{Methodology} \label{sec:method}

@@

\paragraph{Betweenness Centrality}
An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}. By this measure, the "importance" of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points. BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities. Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing. It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex. BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark\cite{ssca2}. Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation. We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
